Testing and Isolation Efficacy: Insights from a Simple Epidemic Model

### 2021-04-29 SIR_manuscript.tex
Meeting with Ali

\newcommand{\Rlogo}{R\xspace}
\newcommand{\percap}{\emph{per capita}\xspace}
\newcommand{\Rnum}{\ensuremath{\mathcal{R}_0}}
\newcommand{\covid}{COVID-19\xspace}
\newcommand{\pro}[1][]{\ensuremath{\frac{\partial #1}{\partial \rho}}}

@@

\begin{figure}[!h] 
\begin{center} 
\includegraphics[scale=2]{pix/sir_comp.pdf}
\caption{\small Flowchart of the SIR (Susceptible-Infectious-Recovered) model, \ref{model}. Here, the disease-based status of a compartment $X$, where $X \in \{S,I,R\}$, is combined with the testing-based status including $X\_u$, $X\_p$, $X\_n$ and $X\_c$, for \emph{untested}, waiting-for-\emph{positive}, waiting-for-\emph{negative}, or \emph{confirmed positive}, respectively. Also,  $\Lambda$ is the force of infection with definition in Eq.~\eqref{Lambda}, $\gamma$ is the recovery rate, $\omega$ is the rate of test return, $F_X$ (defined in Eq. \eqref{F}) and $p_X$ represent the \percap testing rate and the probability of positive tests, respectively, for compartment $X$. For further description of the parameters see \cref{tab:params}.
\label{fig:flowchart}}
\end{center} 
\end{figure}

@@

The Disease-Free Equilibrium, DFE hearafter, for the SIR model, Eqs. \eqref{model}, is given by setting the differential equations to 0 and solving for the unknowns. The DFE is
\begin{equation}
\label{dfe}
S\_n^*= \frac{\rho}{\omega} N_0, \ S\_u^*= N_0-S\_n^*, \text{~and~} I_j=R_j=0 \ \text{for all }j.
\end{equation}
The corresponding \percap testing rate (Eq. \ref{F}) for the infected compartment $I$ at DFE is one of the key analysis parameters and can be simplified as 
\begin{equation}
\label{eq:fi}
\hat F_I = (\omega\rho/(\omega-\rho))w_I/w_S.
\end{equation}

@@

\begin{enumerate}
\item \label{p1:eta} Increasing the isolation parameters for tested and confirmed individuals decreases \Rnum;
$$\partial{\Rnum}/\partial{\theta\_c} \leq 0 \text{~and~} \partial{\Rnum}/\partial{\theta\_w} \leq 0.$$ 
\item \label{p1:rho} A higher testing intensity may increase or decrease $\Rnum$;
$$\partial{\Rnum}/\partial{\rho} \text{~can be positive or negative}.$$
\item \label{p1:omega} A higher rate of test return may increase or decrease \Rnum.
$$\partial{\Rnum}/\partial{\omega} \text{~can be positive or negative}.$$
\end{enumerate}

@@

Our numerical results are shown in Figs. \ref{pan}, where $\rho \in [0,0.01]$, and \ref{pan2}, where $\rho \in [0,0.5]$.
When targeted testing strategy is applied compared to random testing strategy, i.e., comparing panel (b) to (a) of Figs.\,\ref{pan} and \ref{pan2}, it can be inferred that 
(i) \ali{thinking of the intercept of the solid line/contours} the critical \percap testing intensity corresponding to $\Rnum=1$, is lower and thus less testing is required to reduce $\Rnum$, 
(ii) \ali{thinking of the slope contours, having hard time to explain this since there is a non-monotonic effect of omega} sensitivity of $\Rnum$ wrt $1/\omega$; the changes in $\Rnum$ in response to changes in $1/\omega$, is lower.

It is notable that increasing testing intensity $\rho$ as a strategy to reduce $\Rnum$ depends on the value of $\rho$ to begin with, the mean test return time $1/\omega$ and the efficacy of isolation $\theta$. 
In particular, there might be feasible threshold values for $\rho$ and $\omega$ that maximizes the effect of these parameters in reduction of $\Rnum$ depends upon the isolation efficacy.

>>>>

\newcommand{\fref}[1]{Fig.~\ref{#1}}
\newcommand{\Rlogo}{R\xspace}
\newcommand{\percap}{\emph{per capita}\xspace}
\newcommand{\Rnum}{\ensuremath{\mathcal{R}_0}}
\newcommand{\covid}{COVID-19\xspace}
\newcommand{\pro}[1][]{\ensuremath{\frac{\partial #1}{\partial \rho}}}

@@

\begin{figure}
\begin{center} 
\includegraphics[scale=2]{pix/sir_comp.pdf}
\caption{\small Flowchart of the SIR (Susceptible-Infectious-Recovered) model, \ref{model}. Here, the disease-based status of a compartment $X$, where $X \in \{S,I,R\}$, is combined with the testing-based status including $X\_u$, $X\_p$, $X\_n$ and $X\_c$, for \emph{untested}, waiting-for-\emph{positive}, waiting-for-\emph{negative}, or \emph{confirmed positive}, respectively. Also,  $\Lambda$ is the force of infection with definition in Eq.~\eqref{Lambda}, $\gamma$ is the recovery rate, $\omega$ is the rate of test return, $F_X$ (defined in Eq. \eqref{F}) and $p_X$ represent the \percap testing rate and the probability of positive tests, respectively, for compartment $X$. For further description of the parameters see \cref{tab:params}.
\label{fig:flowchart}}
\end{center} 
\end{figure}

@@

The Disease-Free Equilibrium (DFE) for the SIR model, Eqs. \eqref{model}, is found by setting the infected compartments to 0 and solving for the unknowns. The DFE is
\begin{equation}
\label{dfe}
S\_n^*= \frac{\rho}{\omega} N_0, \ S\_u^*= N_0-S\_n^*, \text{~and~} I_j=R_j=0 \ \text{for all }j.
\end{equation}
The corresponding \percap testing rate (Eq. \ref{F}) for the infected compartment $I$ at DFE is one of the key analysis parameters and can be simplified as 
\begin{equation}
\label{eq:fi}
\hat F_I = (\omega\rho/(\omega-\rho))w_I/w_S.
\end{equation}

@@

\begin{enumerate}

\item \label{p1:eta} Increasing the isolation parameters for tested and confirmed individuals decreases \Rnum;

\item \label{p1:rho} Higher testing $\rho$ intensity always decreases $\Rnum$ \textbf{if}: 
%% \begin{enumerate} \item
testing is random; \textbf{or}
%% \item
$\rho$ is small.
%% \end{enumerate}

\item \label{p1:omega} A higher rate of test return always decreases \Rnum, \textbf{if} $\theta_w=0$.

\end{enumerate}

@@

\jd{$w_R$ does not affect \Rnum, and should not be one of the parameters in the figure title. Prepend with English (random testing and weighting testing).}

Numerical results are shown in Fig. \ref{pan}.
Targeted testing (panel b) is more effective at control than random testing (panel a).

When $\theta_w =0$ (top row of each panel), we see that shorter test-waiting times reduce \Rnum\ ($\Delta$ increases as we move to the left in each plot in this row). For non-zero $\Delta$, however, we mostly see the opposite effect: less waiting leads to more transmission. This is because returning negative tests leads people to stop distancing; this applies both to susceptibles, and to infectious people who receive negative test results because they were sampled when susceptible (or because of imperfect test sensitivity, which is not modeled in this figure).

We also see that greater test intensity $\rho$ increases the effectiveness of control $\Delta$ ($\Delta$ increases as we move up in each plot in this Figure. Mathematically speaking, we did find a counter-vailing effect, but this can only be seen when we allow $\rho$ to take unrealistically large values, and only for weighted testing (see \fref{pan2}). The explanation here is that more rapid testing leaves more susceptibles in the ``waiting-for-no'' category at the DFE; these people must then wait for their tests to be returned before they can be tested again, receive a positive test, and become cautious. This effect is usually weak compared to the helpful effects of testing, but if testing is weighted, $\theta_w$ is low, and test returns are slow, t is possible for increasing $\rho$ to reduce $\Delta$ (see upper-right part of upper-right plot of \fref{pan2}b).
